# Regression and model validation

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using  

#### 1. First we will read the data called students2014, explore its structure and dimensions:


```{r read}
learning2014<-read.csv("C:/Users/Elisabet/Desktop/ELI/MASTER'S DEGREE/Open Data Science/GitHub/IODS-project/IODS-project/learning2014.csv", header = TRUE)
```

```{r dim}
dim(learning2014)
```

```{r str}
str(learning2014)
```


We can see the our data has 7 variables (columns), and 166 observations. The data is about the different approaches of learning by students. The variables measured are:

**gender:** It has 2 levels: M for male, F for female

**age:** age of the students. Integral data type variable with a minimum age of 17y.o. and maximum of 55 y.o.

**attitude:** global attitude towards statistics

**deep:** stands for deep approach learning and it is a combination of variables including: seeking meaning, relating ideas and use of evidence.  

**stra:** stands for strategic approach and it is a combination of variables including: organized studying and time management.

**surf:** stands for surface approach learning and it is a combination of variables including: lack of purpose, unrelated memorising and syllabus-boundness.

**points:** the exam points that students obtained. It is an integral data type with a minimum points of 7 and maximum of 33.

The variables attitude, deep, stra and surf are numerical data type with ranging values from 1 to 5.


#### 2. Second, we will do a graphical overview of the data and show the summaries of the variables:

##### First we install the packages and bring them to R studio:

```{r install}

library(ggplot2)

library(GGally)
```

```{r summary}
summary(learning2014)
```



```{r ggpairs}
p <- ggpairs(learning2014, mapping = aes(col=gender), lower=list(combo=wrap("facethist", bins=20)))

p
```